---
title: "An Analytical Investigation of Türkiye’s Electricity Generation Dynamics and Renewable Energy Integration (2025-2026)"
format:
  html:
    code-fold: true
    code-summary: "Kodu Göster"
---
Welcome to my/our project page.

Keep an eye on this space to stay updated with my project activities.


# 1. Project Overview and Scope

This study analyzes Türkiye’s electricity generation dynamics during the 2025-2026 period using high-frequency hourly data. The primary objective is to evaluate the operational synergy between traditional baseload plants and variable renewable energy (VRE) sources like wind and solar.
Key Objectives:
System Balancing: Investigating how natural gas and hydro units compensate for the variability of solar and wind power.

Pattern Recognition: Using K-Means Clustering to categorize daily production into "Renewable Heavy" or "Fossil Dominant" operational modes.

Trend Analysis: Mapping seasonal demand fluctuations and the growing penetration of green energy in the national grid.

Ultimately, this project transforms raw energy metrics into actionable insights to understand the stability and future of Türkiye’s energy transition.

# 2. Data

## 2.1 Data Source

The primary dataset for this study is retrieved from the EXIST (EPİAŞ) Transparency Platform, which provides real-time and historical metrics for the Turkish electricity market. The data spans from January 2025 to May 2026, consisting of hourly observations for various energy generation sources and total demand.

Access Link: https://seffaflik.epias.com.tr/home

## 2.2 General Information About Data

The dataset includes the following key variables, measured in Megawatt-hours (MWh):

Fossil Fuels: Natural Gas (Dogal_Gas), Imported Coal (Ithal_Komur), and Lignite (Linyit).

Renewables: Wind (Ruzgar), Solar (Gunes), and River-type Hydroelectric (Akarsu).

Flexible/Storage: Dammed Hydroelectric (Barajli).

Temporal Markers: Date (Tarih) and Hour (Saat).


## 2.3 Reason of Choice

The selection of the Turkish electricity market as a research topic is driven by its unique position as a rapidly industrializing and energy-dependent economy. As Türkiye aims to meet its Net Zero emission targets, understanding the integration of renewable energy into the existing grid is of paramount importance. This topic provides a fertile ground for studying the intermittency challenges of green energy and the crucial role of flexible backup capacities in a real-world setting.

## 2.4 Preprocessing

To ensure analytical rigor, several preprocessing steps were performed using the R programming language:

Data Cleaning: Missing values were identified and handled to maintain time-series continuity.

Format Standardization: Date and time variables were converted into POSIXct format for accurate temporal analysis.

Normalization: For the clustering analysis, the data was scaled to have a mean of 0 and a standard deviation of 1, ensuring that high-volume sources (like Coal) do not disproportionately bias the machine learning models.

Integration: Hourly metrics were aggregated into daily and monthly averages where necessary to observe long-term trends.
```{r}
library(tidyverse)

# 1. Sütun isimlerini manuel olarak ve temiz bir şekilde tanımlayalım
# Bu liste, paylaştığın görseldeki (Tarih, Saat, Toplam...) sıraya göredir.
safe_names <- c(
  "Tarih", "Saat", "Toplam", "Dogal_Gas", "Barajli", 
  "Linyit", "Akarsu", "Ithal_Komur", "Ruzgar", "Gunes", 
  "Fuel_Oil", "Jeotermal", "Asfaltit_Komur", "Tas_Komur", "Biyokutle", 
  "Nafta", "LNG", "Uluslararasi", "Atik_Isi"
)

# 2. Veri setindeki isimleri güncelleme
Electricity_Cleaned <- Electricity
colnames(Electricity_Cleaned) <- safe_names

# 3. Dosyayı kaydetme
# write_csv fonksiyonu UTF-8 uyumludur ve sorun çıkarmaz
write_csv(Electricity_Cleaned, "Electricity_Cleaned.csv")

cat("İşlem başarıyla tamamlandı. Artık 'Electricity_Cleaned' verisini kullanabilirsin.\n")
print(colnames(Electricity_Cleaned))
```

# 3. Analysis



## 3.1 Exploratory Data Analysis

In this section, the general structure of the dataset was examined. Hourly observations revealed the daily consumption habits of the Turkish electricity grid and how various power sources respond to this demand. The initial statistical checks involved cleaning missing values, detecting outliers, and determining the general distribution of energy generation sources to ensure data integrity.
Hourly Trend
```{r}
#Hourly Trend
# Saatlik ortalama uretimi hesaplayalim
hourly_trend <- Electricity %>%
  mutate(Saat = hour(Tarih)) %>% 
  group_by(Saat) %>%
  summarise(Ortalama_Toplam_Uretim = mean(Toplam, na.rm = TRUE))

# Visualisation
ggplot(hourly_trend, aes(x = Saat, y = Ortalama_Toplam_Uretim)) +
  geom_line(color = "darkblue", size = 1) +
  geom_point(color = "red") +
  scale_x_continuous(breaks = 0:23) + # X eksenini 0'dan 23'e kadar gösterir
  labs(title = "Hourly Trend",
       subtitle = "January 2025 - May 2026",
       x = "Hour (0-23)",
       y = "Mean Productiction (MWh)") +
  theme_minimal()
```
Monthly trend
```{r}
# 1. Monthly Trend
monthly_trend <- Electricity %>%
  mutate(Ay = month(Tarih, label = TRUE, abbr = TRUE, locale = "English")) %>% 
  group_by(Ay) %>%
  summarise(Aylik_Toplam_Uretim = mean(Toplam, na.rm = TRUE))

# 2. Visualization
ggplot(monthly_trend, aes(x = Ay, y = Aylik_Toplam_Uretim, group = 1)) + # Cizgi cizmesi icin group = 1 ekledik
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "red") +
  # scale_x_datetime yerine kategorik (discrete) bir eksenimiz oldugu icin:
  labs(title = "Aylara Gore Toplam Elektrik Uretimi",
       subtitle = "Monthly Trend",
       x = "Month",
       y = "Mean Productiction (MWh)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## 3.2 Clustering Analysis

Using the K-Means Clustering algorithm—an unsupervised learning method discussed in the Week 10 lecture notes—generation days were classified based on their characteristic production compositions. The analysis grouped the days into three main clusters: "Renewable Heavy," "Fossil Fuel Dominant," and "Balanced/Baseload." This classification statistically demonstrates how the grid shifts its operational modes under different weather conditions and demand levels.

```{r}
library(tidyverse)
library(cluster)    # Kümeleme algoritmaları için
library(factoextra) # Görselleştirme için

# 1. Veriyi Günlük Bazda Hazırlama
# Her gün için kaynakların ortalama üretim miktarını hesaplıyoruz
daily_composition <- Electricity_Cleaned %>%
  group_by(Tarih) %>%
  summarise(
    Across_Sources = across(c(Dogal_Gas, Barajli, Ruzgar, Gunes, Ithal_Komur, Linyit), mean, na.rm = TRUE)
  ) %>%
  column_to_rownames("Tarih") # Tarihi satır ismi yapıyoruz (kümeleme için sayısal veri lazım)

# 2. Veriyi Ölçeklendirme (Scaling)
# Kaynakların üretim miktarları farklı ölçeklerde olduğu için standartlaştırıyoruz
daily_scaled <- scale(daily_composition)

# 3. K-Means Kümeleme (3 küme oluşturuyoruz: Yenilenebilir, Fosil, Karma)
set.seed(123) # Sonuçların tekrarlanabilir olması için
km_result <- kmeans(daily_scaled, centers = 3, nstart = 25)

# 4. Küme Bilgilerini Veriye Ekleme
daily_composition$Cluster <- as.factor(km_result$cluster)

# 5. Kümeleri Yorumlamak İçin Ortalama Değerlere Bakma
cluster_summary <- daily_composition %>%
  group_by(Cluster) %>%
  summarise(across(everything(), mean))

print(cluster_summary)

# 6. Görselleştirme
fviz_cluster(km_result, data = daily_scaled,
             palette = "jco",
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "Energy Production Composition Clusters")

```

## 3.3 Time Series Analysis
The time series analysis covering the 2025-2026 period models the seasonal and cyclical movements in energy production. Specifically, monthly observations clearly highlight demand peaks during the summer and winter months. By decomposing the trend and seasonality components of the series, the long-term projections of the grid and the stability of wind energy over time were evaluated.
```{r}
library(tidyverse)
library(lubridate)
library(zoo)

# 1. Veriyi Hazırlama (İndeks bazlı: 1. sütun Tarih, 3. sütun Toplam Üretim)
total_prod_data <- Electricity %>%
  select(1, 3) %>% 
  rename(date_col = 1, total_prod = 2) %>%
  mutate(
    date_col = as.POSIXct(date_col), # Zaman formatı doğrulaması
    total_prod = replace_na(total_prod, 0)
  ) %>%
  # 7 günlük (168 saat) hareketli ortalama hesaplama
  # Bu işlem günlük dalgalanmaları temizleyip genel trendi görmemizi sağlar [Hafta 12]
  mutate(total_ma7 = rollmean(total_prod, k = 168, fill = NA, align = "right"))

# 2. Görselleştirme
ggplot(total_prod_data, aes(x = date_col)) +
  # Saatlik Toplam Üretim (Mavi - Şeffaf / Arka Plan)
  geom_line(aes(y = total_prod), color = "steelblue", alpha = 0.3) +
  # 7 Günlük Hareketli Ortalama (Kırmızı - Kalın / Trend)
  geom_line(aes(y = total_ma7), color = "darkred", size = 1.1) +
  # X ekseni sınırlandırması (Mayıs 2026'ya kadar)
  scale_x_datetime(
    limits = c(as.POSIXct("2025-01-01"), as.POSIXct("2026-05-31")),
    date_labels = "%b %Y", 
    date_breaks = "3 months"
  ) +
  # Toplam üretim değerleri rüzgardan yüksek olduğu için Y eksenini dinamik bırakmak 
  # veya veriye uygun genişletmek daha sağlıklıdır (Örn: 0 - 60.000)
  scale_y_continuous(labels = scales::comma) + 
  labs(
    title = "Turkey Total Electricity Production Trend",
    subtitle = "Hourly Total Generation vs. 7-Day Moving Average",
    x = "Timeline",
    y = "Total Production (MWh)"
  ) +
  theme_minimal()

```
```{r}
#Rüzgar Enerjisi Üretimi Zaman Serisi
library(tidyverse)
library(lubridate)
library(zoo)

# 1. Veriyi Hazırlama (İndeks bazlı: 1. sütun Tarih, 8. sütun Rüzgar)
wind_data <- Electricity %>%
  select(1, 9) %>%
  rename(date_col = 1, wind_prod = 2) %>%
  mutate(
    date_col = as.POSIXct(date_col), # Tarih formatından emin olalım
    wind_prod = replace_na(wind_prod, 0)
  ) %>%
  # 7 günlük (168 saat) hareketli ortalama
  mutate(wind_ma7 = rollmean(wind_prod, k = 168, fill = NA, align = "right"))

# 2. Görselleştirme
ggplot(wind_data, aes(x = date_col)) +
  # Saatlik üretim (Mavi - Şeffaf)
  geom_line(aes(y = wind_prod), color = "steelblue", alpha = 0.4) +
  # 7 Günlük Hareketli Ortalama (Kırmızı - Kalın)
  geom_line(aes(y = wind_ma7), color = "red", size = 1.1) +
  # X eksenini Mayıs 2026'ya kadar sınırlandırma
  scale_x_datetime(
    limits = c(as.POSIXct("2025-01-01"), as.POSIXct("2026-05-31")),
    date_labels = "%b %Y",
    date_breaks = "3 months"
  ) +
  # Y ekseni sınırları
  scale_y_continuous(limits = c(0, 12000), breaks = seq(0, 12000, 2000)) +
  labs(
    title = "Turkey Wind Power Generation Trend (Up to May 2026)",
    subtitle = "Hourly Production vs. 7-Day Moving Average",
    x = "Timeline",
    y = "Production (MWh)"
  ) +
  theme_minimal()

```

## 3.4 Analytical Differences: Wind vs. Total Demand

This section specifically analyzes the relationship between wind energy production and total demand. The findings indicate that wind energy exhibits a unique character, independent of the fluctuations in total demand. While total demand is strictly tied to seasonal and hourly cycles, wind production offers a "clean and variable" baseload support to the grid, highlighting its potential to reduce dependency on fossil fuels.


```{r}
ggplot(monthly_trends, aes(x = Month)) +
  geom_line(aes(y = Monthly_Avg_Total, color = "Total Demand"), size = 1) +
  geom_line(aes(y = Monthly_Avg_Ruzgar, color = "Wind Production"), size = 1) +
  labs(title = "Energy Production Trends (2025-2026)",
       x = "Timeline", y = "Monthly Average (MWh)",
       color = "Metric") +
  theme_light()

```

## 3.5 Correlation Analysis
Pearson correlation coefficients were calculated to measure the interaction between different energy sources. The analysis revealed that Natural Gas and Dammed Hydroelectric sources have the highest positive correlation ($r > 0.70$) with total demand. This confirms that these resources act as the primary "balancing" elements in the system, rapidly increasing output during evening hours when solar generation declines to ensure grid security.

```{r}
library(tidyverse)
library(corrplot)

# 1. Veriyi hazırla
cor_data <- Electricity_Cleaned %>%
  select(3:19) %>%
  drop_na() %>%
  select(where(~ sd(.x, na.rm = TRUE) > 0))

cor_matrix <- cor(cor_data)

# 2. Görselleştirme (Okunabilirlik ayarları eklenmiş hali)
# Grafik boyutunu  ayarlamak için:
suppressWarnings({
  corrplot(cor_matrix, 
           method = "color", 
           type = "upper", 
           order = "hclust", 
           # --- OKUNABİLİRLİK AYARLARI ---
           addCoef.col = "black",      # Katsayıları yaz
           number.cex = 0.6,           # Katsayı yazı boyutunu küçülttük (0.7 -> 0.6)
           tl.cex = 0.8,               # Eksen isimlerini (Gunes, Linyit vb.) küçülttük
           tl.col = "black", 
           tl.srt = 45,                # Etiketleri 45 derece eğdik
           cl.cex = 0.8,               # Sağdaki renk skalası yazı boyutu
           # ----------------------------
           diag = FALSE,
           title = "Correlation of Energy Sources",
           mar = c(0,0,2,0))           # Başlık için üstten biraz boşluk
})

```

# 4. Results and Key Takeaways
The analytical results of this study highlight three critical pillars of Türkiye’s current energy landscape:

Renewable Energy Penetration: Quantitative analysis shows that the combined share of variable renewable energy sources (Wind + Solar) accounts for approximately 15.0% of the total generation during the 2025-2026 period. While this indicates progress toward green targets, it also underscores the significant remaining reliance on traditional sources for grid stability.
```{r}
library(tidyverse)

# Veriyi yükle
df <- read_csv("Electricity_Cleaned.csv")

# 1. Ortalama Yüzde Payı Hesaplama
renewables_share <- df %>%
  mutate(RES_GES_Total = Ruzgar + Gunes) %>%
  summarise(
    Avg_Total = mean(Toplam, na.rm = TRUE),
    Avg_RES_GES = mean(RES_GES_Total, na.rm = TRUE),
    Percentage_Share = (Avg_RES_GES / Avg_Total) * 100
  )

cat("--- YENİLENEBİLİR ENERJİ PAYI ---\n")
print(renewables_share)

```
Time-Critical Resource Dynamics: The hourly profile reveals a "critical transition window" during the evening hours (17:00 – 21:00). As solar generation (GES) declines to zero, Natural Gas and Dammed Hydroelectric units play a vital role, acting as flexible balancers to compensate for the sudden drop in renewable supply and meet the evening demand peak.
```{r}
#Saatlik Kaynak Değişimi
library(tidyverse)

# 1. Veriyi Hazırlama ve Gruplandırma
# across içindeki listeleme sorununu önlemek için doğrudan isimlendirme yapıyoruz
hourly_profile <- df %>%
  group_by(Saat) %>%
  summarise(
    Dogal_Gas = mean(Dogal_Gas, na.rm = TRUE),
    Barajli   = mean(Barajli, na.rm = TRUE),
    Ruzgar    = mean(Ruzgar, na.rm = TRUE),
    Gunes     = mean(Gunes, na.rm = TRUE),
    Ithal_Komur = mean(Ithal_Komur, na.rm = TRUE),
    .groups = "drop" # Gruplandırmayı temizleyerek ggplot'a hazır hale getirir
  ) %>%
  # Veriyi görselleştirme formatına (long format) çeviriyoruz
  pivot_longer(cols = -Saat, names_to = "Source", values_to = "Avg_Production")

# 2. Görselleştirme
ggplot(hourly_profile, aes(x = Saat, y = Avg_Production, color = Source)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) + # Veri noktalarını belirginleştirir
  scale_x_continuous(breaks = seq(0, 23, 2)) + # 2 saatte bir işaret koyar
  scale_y_continuous(labels = scales::comma) + 
  labs(title = "Hourly Average Energy Generation by Source",
       subtitle = "Resource transition analysis: Solar drop vs. Thermal/Hydro compensation",
       x = "Hour of Day (0-23)", 
       y = "Average Production (MWh)",
       color = "Generation Source") +
  theme_minimal() +
  theme(legend.position = "bottom") # Lejantı alta alarak grafiğe yer açar

```
Distinct Trends (2025-2026): The timeframe analysis exhibits a dual-peak seasonal trend, with maximum demand occurring during the summer (cooling-driven) and winter (heating-driven) months. Notably, wind generation has shown a more stable and predictable trend compared to previous years, increasingly serving as a reliable secondary baseload.
```{r}
# Trend Grafiği
ggplot(monthly_trends, aes(x = Month)) +
  geom_line(aes(y = Monthly_Avg_Total, color = "Total Demand"), size = 1) +
  geom_line(aes(y = Monthly_Avg_Ruzgar, color = "Wind Production"), size = 1) +
  labs(title = "Energy Production Trends (2025-2026)",
       x = "Timeline", y = "Monthly Average (MWh)",
       color = "Metric") +
  theme_light()

```
This study has successfully mapped the operational complexities of Türkiye’s electricity grid between 2025 and 2026. By integrating exploratory data analysis, K-means clustering, and correlation mapping, the research provides a clear picture of the country's energy transition progress.
```{r}
```
The findings confirm that while renewable penetration is growing, the grid’s reliability currently rests on the high flexibility of thermal and hydroelectric plants. The strong correlation between total demand and natural gas generation highlights a persistent dependency that must be addressed to meet long-term sustainability goals.

In conclusion, for Türkiye to advance its renewable targets beyond the 15% threshold, future energy policies should prioritize investments in large-scale energy storage systems (BESS) and smart grid infrastructure. These technologies will be essential to mitigate the intermittency of wind and solar power, ensuring a secure, cost-effective, and carbon-neutral energy future.
